Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-03-23 content.tex
3222 words

The goal of \verb;pirouette; is to quantify the importance of a tree prior.
It does so by measuring the inference error made 
for a given reconstructed phylogeny, 
simulated under a (usually novel) diversification model.
We refer to the true model that generated the given tree
as the 'generative tree model' $\mathit{p_{G}}$.
Many tree priors have a parameter setting 
for which they reduce to a standard tree prior. 
For example, a protracted birth-death model [\cite{etienne2012prolonging}] reduces to a standard birth-death model when the speciation-completion rate 
is infinite [i.e. rate $\lambda$ in \cite{etienne2014estimating}, 
eqs.~(2b) and (2c)].
When benchmarking a novel tree prior, 
one will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
While we recommend many replicate simulations 
when assessing a novel tree prior, 
our examples contain only one replicate,
as the goal is to show the workings of \verb;pirouette;,
instead of doing an extensive analysis.
The supplementary materials, however, do show the results
of relicated runs under multiple settings.

\verb;pirouette; is very flexible and allows the user 
to specify a wide variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in Table~\ref{tab:options} and Table~\ref{tab:definitions}.

Although many possible tests can be performed, 
we show the usage of \verb;pirouette; by 
showcasing one single run in high detail,
ending in quantifying the impact a tree prior has in Bayesian inference.

@@

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{workflow4.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by the generative tree model 
    $\mathit{p_{G}}$.
    The phylogeny is converted to an alignment (2a) using the generative alignment model 
    $\mathit{A} = (\mathit{c_{G}}, \mathit{s_{G}})$, composed of a clock and a site model. 
    The user defines one or more experiments.
    For each candidate experiment $\mathit{X_{i}}$ 
    (a combination of inference model $\mathit{I_{i}}$ and condition $\mathit{C_{i}}$),
    if its condition $\mathit{C_{i}}$ is 
    satisfied (which can depend on the alignment), 
    the corresponding inference model $\mathit{I} = \mathit{I_{i}}$ is selected
    to be used in the next step.
    The inference models (3a) of the selected experiments use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using the error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, for each selected inference model a twin pipeline can be run.
    A twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using the twin tree model $\mathit{p_{t}}$, 
    selected among standard diversification models; 
    the default option is the standard birth-death model, with parameters estimated from the original phylogeny.
    A twin alignment (2b) is then simulated from the twin phylogeny 
    using clock model $\mathit{c_{G}}$ and site model $\mathit{s_{G}}$ 
    imported from the generative model. 
    The twin alignment has the same number of mutations as the original alignment.
    The twin pipeline follows the procedure of the main pipeline, 
    resulting in a twin error distribution (5b).
  }
  \label{fig:pipeline}
\end{figure}

We assume the user has a phylogeny simulated with the new diversification model.

The pipeline to assess the error BEAST2 makes in inferring this phylogeny 
then contains the following steps:
\begin{enumerate}
  \item from the given phylogeny an alignment is simulated 
    under a known alignment model $\mathit{A}$;
  \item from this alignment, according to the specified 
    inference conditions $\mathit{C}$, 
    an inference model $\mathit{I}$ is chosen (which may differ from the 
    generative model);
  \item the inference model and the alignment are used 
    to infer a posterior distribution of phylogenies;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to 
    the error measure $\mathit{E}$ specified by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of 
a (DNA) root sequence, a (DNA) mutation rate, a clock model 
and a (DNA) site model.
The root sequence is the DNA sequence of the shared common ancestor,
and is set to four different equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
Supported nucleotide substitution models (part of a DNA
site model) are JC, HKY, TN and GTR. Only the strict
clock model is currently supported in this step.

The second step (Fig.~\ref{fig:pipeline}, 3a)
selects one (or more) inference model(s) $I$ from a set of inference models $I_{1},\dots,I_{n}$. 
We define an experiment $X_{i}$ as the combination of 
an inference model $I_{i}$ and the conditions $C_{i}$ 
to actually use it in the inference step.
For example, we may require that an inference
model (a combination of a tree model, clock model and site model) 
should include the generative/true tree model. 
As a second example, we may require that we have selected a set of 
candidate inference models,
of which only the best should be used in the actual inference.
In the first example, we specified the condition $C_{i}$ that this
generative model should always be run, whereas in the second example,
we specified condition $C_{i}$ that a candidate model should only be run
when it is the best.
The 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood), given the alignment 
simulated in the previous step.
The evidence for an inference model is estimated by nested 
sampling [\cite{russel2019model}], using the \verb;NS; BEAST2 package. 
We note that scripted use of BEAST2 packages is only possible under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences 
using the web interface of \verb;mcbette; [\cite{mcbette}].

@@

The fourth step quantifies the inference error made. 
First the burn-in fraction is removed, i.e. the first phase of the 
Markov chain Monte Carlo (MCMC) run,
which samples an unrepresentative part of state space. 
By default, \verb;pirouette; 
removes the first 10\% of the posterior.
From the remaining posterior \verb;pirouette; 
creates an error distribution, by measuring the difference
between the true tree and each of the posterior 
trees (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The user can specify a function to quantify the differences between
the true and posterior trees. By default, the package uses the nLTT 
statistic (\cite{janzen2015approximate}), which is the absolute difference
between the normalized lineages-through-time plots of two trees.

@@

The first step ($T_1$) of the twinning process creates a tree $\tau_{\mathit{T}}$
with branching times $\Vec{t}_{\mathit{T}}$ while preserving the original
topology $\psi_{\mathit{G}}$:
\begin{align}
  \tau_{\mathit{G}} = (\Vec{t}_{\mathit{G}}, \psi_{\mathit{G}}) 
  \xrightarrow[]{\mathit{T_1}} 
  \tau_{\mathit{T}} = (\Vec{t}_{\mathit{T}}, \psi_{\mathit{G}})
\end{align}
The default option for the diversification model $p_T$ is the standard birth-death model.
It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a birth-death model) 
that maximize this likelihood applied 
to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. We simulate the new branching times using the TESS package (\cite{TESS, hohna2016tess}).

@@

Finally, if the goal is to evaluate BEAST2's performance on a non-standard tree prior, 
the last source of stochasticity comes from phylogenies. 
In fact, a single phylogeny cannot be considered 
as fully representative of the model. 
For this reason multiple phylogenies, 
as well as an equal number of twins, 
must be considered. 
Having the error measure normalized (i.e. comprised in the interval $[0, 1]$), 
it is possible to considerate the aggregated versions of the errors distributions 
across all the runs.
Therefore, if the number of considered phylogenies is high enough, 
the comparison between the main pipeline's aggregated error distribution 
and its twin counterpart leads to a fair evaluation 
of the new tree prior with respect to the baseline error.

@@

We show the usage of \verb;pirouette; on an tree generated 
by a non-standard 
tree model \richel{code at \url{https://github.com/richelbilderbeek/pirouette_example_30}}. 
We use the diversity-dependent tree model \cite{DDD, etienne2011diversity},
which is a birth-death model with a speciation rate that is dependent on the
number of species.

We choose to use a small tree with six taxa, to keep
the calculations short and the figure more readable.
We pick a arbitrary crown age of ten time units.

@@

The first step in \verb;pirouette; is to 
simulate a DNA alignment from the 
given phylogeny, as described in Subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 
mononucleotides each, while the per-nucleotide mutation rate is 
0.1 mutations per unit time.
Together with the crown age of 10 time units, this results
in an expected per-locus mutation probability of 1.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) nucleotide substitution model
and a strict clock model as these are the simplest.
A JC model assumes that mutation rates between nucleotides are equal and 
constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

>>>>

The goal of \verb;pirouette; is to quantify the importance of a tree prior.
It does so by measuring the inference error made 
for a given reconstructed phylogeny, 
simulated under a (usually novel) diversification model.
We refer to the true model that generated the given tree
as the 'generative tree model' $\mathit{p_{G}}$.
We define the birth-death (BD) model (\cite{nee1994reconstructed}) as
the standard tree model, as many (non-standard) tree models 
have a parameter setting for which they reduce to this model. 
For example, a protracted birth-death (PBD) model [\cite{etienne2012prolonging}] 
reduces to a BD model when the speciation-completion rate 
is infinite [i.e. rate $\lambda$ in \cite{etienne2014estimating}, eqs.~(2b) and (2c)].
When benchmarking a novel tree model, 
one will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
While we recommend many replicate simulations 
when assessing a novel tree prior, 
our examples contain only one replicate,
as the goal is to show the workings of \verb;pirouette;,
instead of doing an extensive analysis.
The supplementary materials, however, do show the results
of replicated runs under multiple settings.

\verb;pirouette; allows the user to specify a wide variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in Table~\ref{tab:options} and Table~\ref{tab:definitions}.

@@

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{workflow4.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by the generative tree model 
    $\mathit{p_{G}}$.
    The phylogeny is converted to an alignment (2a) using the generative alignment model 
    $\mathit{A} = (\mathit{c_{G}}, \mathit{s_{G}})$, composed of a clock and a site model. 
    The user defines one or more experiments.
    For each candidate experiment $\mathit{X_{i}}$ 
    (a combination of inference model $\mathit{I_{i}}$ and condition $\mathit{C_{i}}$),
    if its condition $\mathit{C_{i}}$ is 
    satisfied (which can depend on the alignment), 
    the corresponding inference model $\mathit{I} = \mathit{I_{i}}$ is selected
    to be used in the next step.
    The inference models (3a) of the selected experiments use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using the error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, for each selected inference model a twin pipeline can be run.
    A twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using the twin tree model $\mathit{p_{t}}$, 
    selected among standard diversification models; 
    the default option is the standard BD model, 
    with parameters estimated from the original phylogeny.
    A twin alignment (2b) is then simulated from the twin phylogeny 
    using clock model $\mathit{c_{G}}$ and site model $\mathit{s_{G}}$ 
    imported from the generative model. 
    The twin alignment has the same number of mutations as the original alignment.
    The twin pipeline follows the procedure of the main pipeline, 
    resulting in a twin error distribution (5b).
  }
  \label{fig:pipeline}
\end{figure}

The pipeline to assess the error BEAST2 makes in inferring this phylogeny 
then contains the following steps:
\begin{enumerate}
  \item the user supplies one or (ideally) more phylogenies from a 
    new diversification model
  \item from the given phylogeny an alignment is simulated 
    under a known alignment model $\mathit{A}$;
  \item from this alignment, according to the specified 
    inference conditions $\mathit{C}$, 
    an inference model $\mathit{I}$ is chosen (which may differ from the 
    generative model);
  \item the inference model and the alignment are used 
    to infer a posterior distribution of phylogenies;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to 
    the error measure $\mathit{E}$ specified by the user;
\end{enumerate}

The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates an alignment from the given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
For the sake of clarity, here we will assume the alignment consists out
of DNA sequences, but other heritable material such as amino acids also works.
The user must specify a root sequence, a mutation rate and a site model.
The root sequence is the DNA sequence of the shared common ancestor,
and is set to four different equally-sized mononucleotide blocks by default, 
which helps interpreting the resulting alignment.
Supported nucleotide substitution models (part of a DNA
site model) are JC, HKY, TN and GTR (see \ref{tab:options} for
the meaning of these abbreviatons).
JC is the default nucleotide subsitution mode (NSM),
in which the nucleotide substition rates between all
nucleotides are equal and constant over time.
Note that the user cannot specify a clock model, as only the strict clock model 
is implemented and thus always assumed.

The second step (Fig.~\ref{fig:pipeline}, 3a)
selects one (or more) inference model(s) $I$ from a set of inference models $I_{1},\dots,I_{n}$. 
We define an experiment $X_{i}$ as the combination of 
an inference model $I_{i}$ and the conditions $C_{i}$ 
to actually use it in the inference step.
For example, we may require that an inference
model (a combination of a tree model, clock model and site model) 
should include the generative/true tree model. 
As a second example, we may require that we have selected a set of 
candidate inference models,
of which only the best should be used in the actual inference.
In the first example, we specified the condition $C_{i}$ that this
generative model should always be run, whereas in the second example,
we specified condition $C_{i}$ that a candidate model should only be run
when it is the best.
The 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood), based on the alignment 
simulated in the previous step.
The evidence for an inference model is estimated by nested 
sampling [\cite{russel2019model}], using the \verb;NS; BEAST2 package. 
We note that scripted use of BEAST2 packages is only possible under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences 
using the web interface of \verb;mcbette; [\cite{mcbette}].

@@

The fourth step quantifies the inference error made. 
First the burn-in fraction is removed, i.e. the first phase of the 
Markov chain Monte Carlo (MCMC) run,
which samples an unrepresentative part of state space. 
By default, \verb;pirouette; 
removes the first 10\% of the posterior.
From the remaining posterior \verb;pirouette; 
creates an error distribution, by measuring the difference
between the true tree and each of the posterior 
trees (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The user can specify a function to quantify the differences between
the true and posterior trees. By default, the package uses the nLTT 
statistic (\cite{janzen2015approximate}), which is the absolute difference
between the normalized lineages-through-time plots of two trees.
The nLTT statistic is chosen, as it can operate on any two trees (regardless
of their crown ages and number of taxa) and its results have a clear range
from zero to one. This normalized result makes it possible to compare trees 
from a distribution of trees from any tree model.

@@

The first step ($T_1$) of the twinning process creates a tree $\tau_{\mathit{T}}$
with branching times $\Vec{t}_{\mathit{T}}$ while preserving the original
topology $\psi_{\mathit{G}}$:
\begin{align}
  \tau_{\mathit{G}} = (\Vec{t}_{\mathit{G}}, \psi_{\mathit{G}}) 
  \xrightarrow[]{\mathit{T_1}} 
  \tau_{\mathit{T}} = (\Vec{t}_{\mathit{T}}, \psi_{\mathit{G}})
\end{align}
The default option for the diversification model $p_T$ is the standard BD model.
It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a BD model) 
that maximize this likelihood applied 
to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. 
We simulate the new branching times using the TESS package (\cite{TESS, hohna2016tess}).

@@

Finally, if the goal is to evaluate BEAST2's performance on a non-standard tree prior, 
one must also consider the last source of stochasticity: 
the different phylogenies a tree model generates.
A single phylogeny cannot be considered as fully representative of the model. 
For this reason multiple phylogenies, as well as an equal number of twins, must be considered. 
If the number of considered phylogenies is high enough, 
the comparison between the main pipeline's aggregated error distribution 
and its twin counterpart leads to a fair evaluation 
of the new tree prior with respect to the baseline error.

@@

We show the usage of \verb;pirouette; on an tree generated 
by the non-standard diversity-dependent tree model \cite{DDD, etienne2011diversity},
which is a birth-death model with a speciation rate that is dependent on the number of species.

We use a small tree with six taxa, to keep the figure more readable,
with an arbitrary crown age of ten time units.

@@

The first step in \verb;pirouette; is to 
simulate a DNA alignment from the 
given phylogeny, as described in Subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 
mononucleotides each, while the per-nucleotide mutation rate is 
0.1 mutations per unit time.
Together with the crown age of 10 time units, this results
in an expected per-locus mutation probability of 1.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) NSM
and a strict clock model as these are the simplest.
A JC model assumes that mutation rates between nucleotides are equal and 
constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.
